Dynamics of Laterally Propagating Flames in X-ray Bursts

### 2019-11-24 xrb/paper.tex
Edited caption for plot showing AMR grid

\author[0000-0001-8092-1974]{Weiqun Zhang}
\affiliation{Lawrence Berkeley National Laboratory, Berkeley, CA}

\correspondingauthor{Michael Zingale}
\email{Michael.Zingale@stonybrook.edu}

@@

\cite{cavecchi:2013,art-2015-cavecchi-etal,art-2016-cavecchi-etal,Cavecchi2017,Cavecchi2019} \MarginPar{need a summary}

@@

All simulations are performed with the \castro\ hydrodynamics
code~(\citealt{castro}; see also \citealt{astronum:2017} for a recent
description).  We evolve the system of fully compressible Euler
equations for reacting flow:
\begin{eqnarray}
\frac{\partial( \rho X_k)}{\partial t} &=& - \nabla\cdot (\rho \Ub X_k) + \rho \omegadot_k \\
%
\frac{\partial (\rho \Ub)}{\partial t} &=& - \nabla\cdot (\rho \Ub \Ub) - \nabla p +
    \rho \gb \nonumber \\
  &&-2 \rho \Omegab\times \Ub - \Omegab \times (\Omegab \times \rb) \\
%
\frac{\partial (\rho E)}{\partial t} &=& - \nabla \cdot (\rho \Ub E + p\Ub) +
    \nabla \cdot \kth \nabla T + \rho \epsdot +\nonumber\\
  && \rho \Ub \cdot \gb - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho |\Omegab|^2 (\Ub \cdot \rb)
\end{eqnarray}
Here, $\rho$ is the mass density, $\Ub$ is the velocity, $p$ is the pressure and $E$ is
the specific total energy, which is related to the specific internal
energy as $e = E - |\Ub|^2/2$.  Species are described by mass
fractions, $X_k$ (such that $\sum_k X_k = 1$), and creation rates,
$\omegadot$, and are related to the total specific energy generation
rate, $\epsdot$.  The total mass conservation,
\begin{equation}
\frac{\partial \rho}{\partial t} = - \nabla\cdot (\rho \Ub)
\end{equation}
implies $\sum_k \omegadot_k = 0$.  \castro\ uses an unsplit piecewise
parabolic method (PPM) with characteristic tracing for solving the
hydrodynamics~\citep{ppm,millercolella:2002}, generalized to an
arbitrary equation of state~\citep{zingalekatz}.  Reactions are
incorporated via Strang splitting~\citep{strang:1968}, giving a method
that is overall second-order accurate in space and time.
\castro\ uses the \amrex\ adaptive mesh refinement
library~\citep{amrex_joss} to manage a hierarchy of grids at different
resolutions.

@@

\begin{figure}[t]
	\plotone{grids.pdf}
	\caption{\label{fig:grids} Slice through a simulation showing the four-level
		grid structure. Note that the boxes shown are not the simulation zones, which
		span $10~\mathrm{cm}$ at the finest level, but subdomains containing
		approximately equal numbers of zones that are each managed by a separate
		MPI process. A coarse base grid extending to the upper boundary is drawn
		in white, with the green, blue, and black grids showing the jumps in
		refinement needed to fully resolve the fuel layer and underlying neutron star.}
\end{figure}

@@

The speed and thickness of a laminar helium flame are determined by the
energy generation rate and conductivity, and scale roughly as 
\begin{equation}
v_f \approx \sqrt{k_\mathrm{th} \dot{\epsilon}} \qquad
\lambda_f \approx \sqrt{\frac{k_\mathrm{th}}{\dot{\epsilon}}}
\end{equation}
\citep{orourke:1979,khokhlov:1993}.
At the densities we consider in this simulation, the pure He flame
speed is quite slow and would require a prohibitively long integration
time to see significant evolution of the burning.  We therefore
consider boosted flames in this first paper, to accelerate the burning
and allow us to understand the qualitative effects of laterally
propagating flames.  To boost the flame while keeping the thickness
the same, we can multiply both the burning rate and the conductivity
by the same factor.  For our standard calculations, we choose 10 for
each, to give a 10$\times$ faster flame speed.  We call this the ``10/10''
flame.  We will also do a simulation with the reactions and
conductivity both boosted by 5, the ``5/5'' flame.

@@

\begin{figure*}[t]
\plottwo{flame.pdf}{speed.pdf}
\caption{\label{fig:flame} Time-evolution of the 10$\times$ boosted 1D
	laminar flame. The left plot shows temperature and nuclear energy
	generation profiles at 11 different time points, while the right plot
	shows flame propagation speed and flame thickness as functions of time.}
\end{figure*}

@@

\begin{deluxetable}{lcc}
	\tablecaption{\label{table:flame_speeds}Turbulent and laminar flame speeds.}
	\tablehead{\colhead{run} & \colhead{$s_T$ (km s$^{-1}$)} & \colhead{$s_L$ (km s$^{-1}$)}}
	\startdata
	10/10 boost & $9.18 \pm 0.03$  & $1.06 \pm 0.01$ \\
	5/5 boost & $4.00 \pm 0.01$ & $0.56 \pm 0.01$ \\
	iso7 network & $7.56 \pm 0.02$ & $-$ \\
	lowres & $9.33 \pm 0.04$ & $-$ \\
	1000 Hz & $18.6 \pm 0.16$ & $-$ \\
	\enddata
	%
\end{deluxetable}

@@

Creating the model involves specifying $T_\star$, $T_\mathrm{hi}$,
$T_\mathrm{lo}$, $\rho_\mathrm{int}$, $H_\star$,
$\delta_\mathrm{atm}$, ${X_k}_\star$, and ${X_k}_\mathrm{atm}$.  We
then integrate outwards from the base of the ramp region ($y =
H_\star$), enforcing the discrete form of hydrostatic equilibrium,
Eq.~\ref{eq:hse}.  Integrating upwards, we find $p_i$ and
$\rho_i$ using a Newton solve together with the equation of state, with either the temperature
specified, $T_i = T(p_i, \rho_i, \{X_k\}_i)$ (in the isothermal, ramp,
and buffer layers) or constant entropy, $s_i = s(p_i, \rho_i,
\{X_k\}_i)$, in the fuel layer.  This follows the procedures described in \citet{ppm-hse}.  We
use a constant temperature for all $y < H_\star +
3\delta_\mathrm{atm}$.  Above this, we switch to an isentropic atmosphere until the
temperature drops to a floor value, $T_\mathrm{lo}$, at which point we
again keep the temperature constant.  The integration of the
atmosphere continues until the density falls to a low density cutoff,
$\rho_\mathrm{cutoff}$.  The material above this height is taken to
have constant density and temperature.

@@

Table \ref{table:flame_speeds} gives the flame speed measured in each
multidimensional simulation, as well as laminar speeds obtained from
corresponding 1D runs. The 2D flames propagate at speeds about an
order of magnitude faster than their 1D counterparts. The increase in
flame speed is likely a product of the larger flame surface area and hydrodynamical effects such as
turbulence, wrinkling, and convective cycles, which bring cooler fuel
from ahead of the front into the hottest part of the burning region.

@@

{\color{red} Should say qualitatively what the plot shows}

@@

Baroclinicity is calculated as
\begin{equation}
    \boldsymbol{\psi} = \frac{1}{\rho^2} \nablab p \times \nablab \rho,
\end{equation}
and shows the misalignment of the local density and pressure gradients
\citep{Malone2014a}. As we are considering a 2d \MarginPar{We've used at least three different ways of writing 2-d/2D/2d so far. These should be changed so we used one style throughout.}system, the component of the
baroclinicity we consider here (out of the plane) reflects the misalignment
of the fields in the plane of the simulation. Figure~\ref{fig:baroclinicity}
shows that the baroclinicity peaks along the flame front. This baroclinicity
generates vorticity, which in turn entrains material along the surface of the
flame.

@@

To further demonstrate the relationships between different properties of the flame, in Figure~\ref{fig:phase_plots} we present some phase plots of the energy generation rate. The phase plot of the
energy generation rate as a function of
the $x$- and $y$-velocities shows that energy is preferentially generated in
regions with negative $x$-velocity. This most likely corresponds to the burning along the underside of the flame front, where fuel has been entrained along the surface of the flame and so is moving in the opposite direction to the direction the flame is propagating in. This is further supported by the phase plot of the energy generation rate as a function of the $x$-velocity
and the density, which shows that the energy generation rate peaks at
$\rho \sim 3 \times 10^5~\mathrm{g/cm}^3$, at the base of the flame.

@@

In order to satisfy these requirements whilst allowing the simulations to be computationally feasible, we used several approximations: a higher than normal rotation rate, a boosted flame, a simplified reaction network, and a 2.5D model for the flow. The first two of these were used to reduce the model's spatial and temporal scales. Using the simulation framework developed here, these both can be relaxed in the future, at the cost of more computer time. The same goes for 2.5D
vs.\ full 3D---the only difference is computer time, and our future
calculations will explore the 3D evolution and compare to the 2.5D simulations presented here. In particular,
in 3D we will be able to explore shear instabilities at the flame front.
Larger networks are a straightforward change, and already supported in
\castro\ using the \pynucastro\ framework~\citep{pynucastro} and JINA
ReacLib rate database~\citep{reaclib}.

@@

\acknowledgements \castro\ is open-source and freely available at
\url{http://github.com/AMReX-Astro/Castro}.  The problem setups used
here are available in the git repo as {\tt flame} and {\tt
  flame\_wave}.  The work at Stony Brook was supported by DOE/Office
of Nuclear Physics grant DE-FG02-87ER40317.  This research used
resources of the National Energy Research Scientific Computing Center,
a DOE Office of Science User Facility supported by the Office of
Science of the U.~S.\ Department of Energy under Contract
No.\ DE-AC02-05CH11231.  This research used resources of the Oak Ridge
Leadership Computing Facility at the Oak Ridge National Laboratory,
which is supported by the Office of Science of the U.S. Department of
Energy under Contract No. DE-AC05-00OR22725, awarded through the DOE
INCITE program.  This research has made use of NASA's Astrophysics
Data System Bibliographic Services.

>>>>

\correspondingauthor{Kiran Eiden}
\email{kiran.eiden@stonybrook.edu}

@@

The first vertically resolved simulations of lateral deflagrations
were obtained by \citet{cavecchi:2013} who showed how the Coriolis
force creates a geometrical configurations that increases the flame
speed set by conduction by a factor $\sim ~ L_r / H$, where $L_r$ is the
Rossby radius and $H$ is the scale height of the burning layer. The
effect on flame propagation of changing Coriolis confinement across
the surface was explored in \citet{art-2015-cavecchi-etal}, while \citet{art-2016-cavecchi-etal}
showed how the magnetic field tension, opposing the Coriolis
force, can either speed up or slow down the flame by changing the
flame front orizontal extent. Finally, \citet{Cavecchi2019}
studied the effects in 3D of the baroclinic instability at the flame
front, measuring flames up to 10 times faster than in the 2D case.

@@

All simulations are performed with the \castro\ hydrodynamics
code~(\citealt{castro}; see also \citealt{astronum:2017} for a recent
description).  We evolve the system of fully compressible Euler
equations for reacting flow:
\begin{eqnarray}
\frac{\partial( \rho X_k)}{\partial t} &=& - \nabla\cdot (\rho \Ub X_k) + \rho \omegadot_k \\
%
\frac{\partial (\rho \Ub)}{\partial t} &=& - \nabla\cdot (\rho \Ub \Ub) - \nabla p +
    \rho \gb \nonumber \\
  &&-2 \rho \Omegab\times \Ub - \rho \Omegab \times (\Omegab \times \rb) \\
%
\frac{\partial (\rho E)}{\partial t} &=& - \nabla \cdot (\rho \Ub E + p\Ub) +
    \nabla \cdot \kth \nabla T + \rho \epsdot +\nonumber\\
  && \rho \Ub \cdot \gb - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho |\Omegab|^2 (\Ub \cdot \rb)
\end{eqnarray}
Here, $\rho$ is the mass density, $\Ub$ is the velocity, $p$ is the pressure and $E$ is
the specific total energy, which is related to the specific internal
energy as $e = E - |\Ub|^2/2$.  Species are described by mass
fractions, $X_k$ (such that $\sum_k X_k = 1$), and creation rates,
$\omegadot$, and are related to the total specific energy generation
rate, $\epsdot$.  The total mass conservation,
\begin{equation}
\frac{\partial \rho}{\partial t} = - \nabla\cdot (\rho \Ub)
\end{equation}
implies $\sum_k \omegadot_k = 0$.  \castro\ uses an unsplit piecewise
parabolic method (PPM) with characteristic tracing for solving the
hydrodynamics~\citep{ppm,millercolella:2002}, generalized to an
arbitrary equation of state~\citep{zingalekatz}.  Reactions are
incorporated via Strang splitting~\citep{strang:1968}, giving a method
that is overall second-order accurate in space and time.
\castro\ uses the \amrex\ adaptive mesh refinement
library~\citep{amrex_joss} to manage a hierarchy of grids at different
resolutions.

@@

\begin{figure}[t]
	\plotone{grids.pdf}
	\caption{\label{fig:grids}A slice through part of a 2D simulation showing the four-level
		grid structure. Note that the boxes shown are not the simulation zones, which span
		$10~\mathrm{cm}$ at the finest level, but subdomains containing approximately equal
		numbers of zones that are each managed by a separate MPI process. A coarse base grid
		extending to the upper boundary is drawn in white, with the green, blue, and black
		grids showing the jumps in refinement needed to fully resolve the fuel layer and
		underlying neutron star.}
\end{figure}

@@

The speed and thickness of a laminar helium flame are determined by the
energy generation rate and conductivity, and scale roughly as 
\begin{equation}
\label{eq:flame_scaling}
v_f \approx \sqrt{\kth \dot{\epsilon}} \qquad
\lambda_f \approx \sqrt{\frac{\kth}{\dot{\epsilon}}}
\end{equation}
\citep{orourke:1979,khokhlov:1993}.
At the densities we consider in this simulation, the pure He flame
speed is quite slow and would require a prohibitively long integration
time to see significant evolution of the burning.  We therefore
consider boosted flames in this first paper, to accelerate the burning
and allow us to understand the qualitative effects of laterally
propagating flames.  To boost the flame while keeping the thickness
the same, we can multiply both the burning rate and the conductivity
by the same factor.  For our standard calculations, we choose 10 for
each, to give a 10$\times$ faster flame speed.  We call this the ``10/10''
flame.  We will also do a simulation with the reactions and
conductivity both boosted by 5, the ``5/5'' flame.

@@

\begin{figure*}[t]
\plottwo{flame.pdf}{speed.pdf}
\caption{\label{fig:flame} Time-evolution of the 10$\times$ boosted 1D
	laminar flame. The left plot shows temperature and nuclear energy
	generation profiles at 11 different times, while the right plot
	shows flame propagation speed and flame thickness as functions of time.}
\end{figure*}

@@

\begin{deluxetable}{lc}
	\tablecaption{\label{table:flame_speeds} Laminar flame speeds.}
	\tablehead{\colhead{run} & \colhead{$s_L$ (km s$^{-1}$)}}
	\startdata
	10/10 boost & $1.06 \pm 0.01$ \\
	5/5 boost & $0.56 \pm 0.01$ \\
	\enddata
	%
\end{deluxetable}

@@

Creating the model involves specifying $T_\star$, $T_\mathrm{hi}$,
$T_\mathrm{lo}$, $\rho_\mathrm{int}$, $H_\star$,
$\delta_\mathrm{atm}$, ${X_k}_\star$, and ${X_k}_\mathrm{atm}$.  We
then integrate outwards from the base of the ramp region ($y =
H_\star$), enforcing the discrete form of hydrostatic equilibrium,
Eq.~\ref{eq:hse}.  Integrating upwards, we find $p_i$ and
$\rho_i$ using a Newton-Raphson solver together with the equation of state, with either the temperature
specified, $T_i = T(p_i, \rho_i, \{X_k\}_i)$ (in the isothermal, ramp,
and buffer layers) or constant entropy, $s_i = s(p_i, \rho_i,
\{X_k\}_i)$, in the fuel layer.  This follows the procedures described in \citet{ppm-hse}.  We
use a constant temperature for all $y < H_\star +
3\delta_\mathrm{atm}$.  Above this, we switch to an isentropic atmosphere until the
temperature drops to a floor value, $T_\mathrm{lo}$, at which point we
again keep the temperature constant.  The integration of the
atmosphere continues until the density falls to a low density cutoff,
$\rho_\mathrm{cutoff}$.  The material above this height is taken to
have constant density and temperature.

@@

\begin{deluxetable}{lc}
	\tablecaption{\label{table:flame_speeds_multid} Flame speeds measured from 2D calculations.}
	\tablehead{\colhead{run} & \colhead{$s_\mathrm{front}$ (km s$^{-1}$)}} 
	\startdata
	10/10 boost & $9.18 \pm 0.03$ \\
	5/5 boost & $4.00 \pm 0.01$ \\
	iso7 network & $7.56 \pm 0.02$ \\
	lowres & $9.33 \pm 0.04$ \\
	1000 Hz & $18.6 \pm 0.16$ \\
	\enddata
	%
\end{deluxetable}

Table \ref{table:flame_speeds_multid} gives the flame speed measured in each
multidimensional simulation, as well as laminar speeds obtained from
corresponding 1D runs. The 2D flames propagate at speeds about an
order of magnitude faster than their 1D counterparts. The increase in
flame speed is likely a product of the larger flame surface area and hydrodynamical effects such as
turbulence, wrinkling, and convective cycles, which bring cooler fuel
from ahead of the front into the hottest part of the burning region.

@@

As we see in Figure~\ref{fig:flame_speed}, all of the various
approximations behave as we would expect.  The 1000~Hz run goes 2
times faster than the 2000~Hz run, as expected by the reduced rossby
radius \citep{cavecchi:2013}).  The 5/5 boost run is half the speed of
the 10/10 boost run, also as expected, consistent with the 1D laminar
flames and behaving as predicted by Eq.~\ref{eq:flame_scaling}.  This
confirms that the role of the Coriolis force is really to boost the
laminar flame by the geometrical/hydrodynamical effect expected ($v_f
= v_l * L_r / H$), similarly to what the 1000 Hz simulation
said.

%Seeing the proofs both changing the Coriolis force AND the
%laminar flame properties I think is very cool.

@@

Baroclinicity is calculated as
\begin{equation}
    \boldsymbol{\psi} = \frac{1}{\rho^2} \nablab p \times \nablab \rho,
\end{equation}
and shows the misalignment of the local density and pressure gradients
\citep{Malone2014a}. As we are considering a 2d \MarginPar{We've used
  at least three different ways of writing 2-d/2D/2d so far. These
  should be changed so we used one style throughout.}system, the
component of the baroclinicity we consider here (out of the plane)
reflects the misalignment of the fields in the plane of the
simulation. Figure~\ref{fig:baroclinicity} shows that the
baroclinicity peaks along the flame front. This baroclinicity
generates vorticity, which in turn entrains material along the surface
of the flame (see also \citet{cavecchi:2013}). This same vorticity,
in 3D, should perturb the flame front and further increase flame speed
\citep{Cavecchi2019}.

@@

To further demonstrate the relationships between different properties
of the flame, in Figure~\ref{fig:phase_plots} we present some phase
plots of the energy generation rate. The phase plot of the energy
generation rate as a function of the $x$- and $y$-velocities shows
that energy is preferentially generated in regions with negative
$x$-velocity. This most likely corresponds to the burning along the
underside of the flame front, where fuel has been entrained along the
surface of the flame and so is moving in the opposite direction to the
direction the flame is propagating in. This is further supported by
the phase plot of the energy generation rate as a function of the
$x$-velocity and the density, which shows that the energy generation
rate peaks at $\rho \sim 3 \times 10^5~\mathrm{g/cm}^3$, at the base
of the flame.  The energy generation rate also depends strongly on density.

@@

In order to satisfy these requirements whilst allowing the simulations to be computationally feasible, we used several approximations: a higher than normal rotation rate, a boosted flame, a simplified reaction network, and a 2.5D model for the flow. The first two of these were used to reduce the model's spatial and temporal scales. Using the simulation framework developed here, these both can be relaxed in the future, at the cost of more computer time. The same goes for 2.5D
vs.\ full 3D---the only difference is computer time, and our future
calculations will explore the 3D evolution and compare to the 2.5D simulations presented here. In particular,
in 3D we will be able to explore shear instabilities at the flame front.  We can also capture the 
baroclinic instability \citep{Cavecchi2019} and the competition between it and shear.
Larger networks are a straightforward change, and already supported in
\castro\ using the \pynucastro\ framework~\citep{pynucastro} and JINA
ReacLib rate database~\citep{reaclib}.

@@

\acknowledgements \castro\ is open-source and freely available at
\url{http://github.com/AMReX-Astro/Castro}.  The problem setups used
here are available in the git repo as {\tt flame} and {\tt
  flame\_wave}.  The work at Stony Brook was supported by DOE/Office
of Nuclear Physics grant DE-FG02-87ER40317.  YC was supported by the
European Union Horizon 2020 research and innovation program under
the Marie Sklodowska-Curie Global Fellowship grant agreement No.\
703916.  This research used resources of the National Energy Research
Scientific Computing Center, a DOE Office of Science User Facility
supported by the Office of Science of the U.~S.\ Department of Energy
under Contract No.\ DE-AC02-05CH11231.  This research used resources
of the Oak Ridge Leadership Computing Facility at the Oak Ridge
National Laboratory, which is supported by the Office of Science of
the U.S. Department of Energy under Contract No. DE-AC05-00OR22725,
awarded through the DOE INCITE program.  This research has made use of
NASA's Astrophysics Data System Bibliographic Services.
